A Reproducible Workflow with RMarkdown, Git, Make & Docker

### 2019-08-14 rr-flow.Rmd
check spelling & grammar

Reproducible Research has become the gold standard for scientific work in recent years. The following tutorial aims to outline a workflow that does not only garanties a sane standard of reproducibility but also decreases the burden of maintaining software for the researcher and his colleages. The authors beflief that integrating tools from software engineering in the process of scientific reporting to enhance repdoducibility is crucial and give guidelines for the practitioner to use them. These tools include software for literate programming, version control, dependency tracking and containerization.

Literate Programming
: Interwoven text and computer code that together form a human readable document.
: Example: One document that includes text that descibes a statistical test, code that is used to execute the statistical and the output of the code in the same document.

Version Control
: The management of change of documents/code.
: Example: Changes from the manuscript text upon request by the reviewer are clearly identifiable.

Dependency Tracking
: How different parts of a manuscript/analysis do relate to each other and how these dependences are resolved.
: Example: The manuscript depends on the statistical analysis which depends on the data.

Containerization
: A virtual computational environment that does encapsulate all needed software. More formally called operating-system-level virtualization.
: Example: A statistical analysis depends on the statistical software that is used and they are bundled together.

@@

```{r, eval=FALSE}
# BAD
iris <- read.csv("/home/aaron/Documents/reproducible-research/data/iris.csv")
# GOOD
iris <- read.csv("data/iris.csv")
# GOOD
iris <- read.csv(here("data", "iris.csv"))
```

This folder, where everything resides that you need for an analysis, will be refered to as a Projekt. Working with Projects is especially easy with RStudio an integrated development environment (IDE) for R.

Reproducibility at a basic level assumes that results are the same, if neither script nor data has changed. It is often not trivial to find out wether anything has changed and if so to "go back in time". `git` enables you to do both. A good mental model for `git` is that it takes a sequenze of snapshot of all files it is supposed to track. These snapshots are reffered to as "commits". Each commit has a unique identifier (a hash) and each commits represents a complete copy of the state of the files at the point when they where added to a commit. Going back to one state is as easy has finding out the hash of the commit. This can be made even easier by tagging specific commits e.g. as "submission", "preprint", "publication". This collection of commits is called "repository", which is almost the same as your Project.

A typicall workflow looks like that:

@@

So to just keep track of all changes you only need to use `git add` & `git commit` or `git commit -a` to conviently do both. These commands need to be executed in the `terminal`, which you can access from within Rstudio (`Shift + Alt + R`), but if you use Git with Rstudio, you also get a graphical user interface.

@@

Now you can check everything that has happend (`git log`) and look at the exact state by suppling the hash or the first few digits of it to `git checkout`:

@@

Git also makes it especially easy to share and collaborate on a project with another researcher. A popular service for sharing via git is [Github](https://github.com). Just sharing Git repositories with the public is allways free, privat repositories (only visible to persons you invite) are [free for researcher](https://help.github.com/en/articles/applying-for-an-educator-or-researcher-discount). After creating a user account, you can create a new repository where Github advises you how to upload you repository from the terminal:

@@

Each update can then be uploaded with `git push` or the green up arrow in the git pane. To download this git repo to another computer just type:

@@

Git & Github can do even more to help you to collaborate with your fellow researcher, but this is beyond the already broad scope of this tutorial.

>>>>

Reproducible Research has become the gold standard for scientific work in recent years. The following tutorial aims to outline a workflow that does guarantees not only a reasonable standard of reproducibility but also decreases the burden of maintaining software for the researcher and his colleagues. The authors' believe that integrating tools from software engineering in the process of scientific reporting to enhance reproducibility is crucial and give guidelines for the practitioner to use them. These tools include software for literate programming, version control, dependency tracking, and containerization.

Literate Programming
: Interwoven text and computer code that together forms a human-readable document.
: Example: One document that includes text that describes a statistical test, code that is used to execute the statistical and the output of the code in the same document.
Version Control
: The management of change of documents/code.
: Example: Changes from the manuscript text upon request by the reviewer are identifiable.

Dependency Tracking
: How different parts of a manuscript/analysis relate and how these dependencies resolve via execution of code.
: Example: The manuscript depends on the statistical analysis, which depends on the data.

Containerization
: A virtual computational environment that does encapsulate all needed software. More formally called operating-system-level virtualization.
: Example: A statistical analysis depends on the used statistical software and a virtual machine bundles them together.

@@

```{r, eval=FALSE}
# BAD
iris <- read.csv("/home/aaron/Documents/reproducible-research/data/iris.csv")
# GOOD
iris <- read.csv("data/iris.csv")
# BETTER
iris <- read.csv(here("data", "iris.csv"))
```

This folder, where everything resides that you need for analysis, is referred to as a Projekt. Working with Projects is exceptionally comfortable with RStudio an integrated development environment (IDE) for R.

Reproducibility at a basic level assumes that results are the same if neither script nor data have changed. It is often not trivial to find out whether anything has changed and if so to "go back in time." `git` enables you to do both. A good mental model for `git` is that it takes a sequence of snapshots of all files it is supposed to track. In the language of git, these snapshots are "commits".  A commit represents a complete copy of the state of the files when added to a commit. Each commit has a unique identifier (a hash). Going back to one state is as easy as finding the hash of the commit. This can be made even easier by tagging specific commits e.g., as "submission," "preprint," "publication." This collection of commits is called "repository," which is almost the same as your Project.

A typical git workflow looks like that:

@@

So to keep track of all changes, you only need to use `git add` & `git commit` or `git commit -a` to do both. These commands need to be executed in the `terminal`, which you can access from within Rstudio (`Shift + Alt + R`), but if you use Git with Rstudio, you also get a graphical user interface and quite likely you won't need the terminal more then once per project.

@@

Now you can check everything that has happened (`git log`) and look at the exact state by supplying the id of the commit or the first few digits of it to `git checkout`:

@@

Git also makes it especially easy to share and collaborate on a project with another researcher. A popular service for sharing via git is [Github](https://github.com). Just sharing Git repositories with the public is always free, private repositories (only visible to persons you invite) are [free for researcher](https://help.github.com/en/articles/applying-for-an-educator-or-researcher-discount). After creating a user account, you can create a new repository where Github advises you how to upload your repository from the terminal e.g., for this repository:

@@

`git push` or the green up arrow in the git pane uploads local updates. To download this git repo to another computer type into the terminal:

@@

Git & Github can do even more to help you to collaborate with your fellow researcher, but this is beyond the already broad scope of this tutorial. Another benefit of using Git & Github is that experimentation is highly encouraged since you can go back to any state quickly and even when you lose access to the file on your computer, everything is backed up on the Github servers.

Even when you have obtained a version of the project, and you can ensure that this version is unchanged, you may not know how to reproduce the results, because it is unclear what has to be run in which order. This job gets comfortable with `make`, since it allows you to create (computational) recipes too (re-)create files.

A `Makefile` obeys the following scheme:

```
target: dependency1 dependency2
  command-to-create-target
```

A typical `Makefile` might look like:

```
analysis.pdf: data/clean.csv
  Rscript -e 'rmarkdown::render("analysis.Rmd")'

data/clean.csv: R/clean.R data/iris.csv
  Rscript -e 'source("R/clean.R")'
```

The first line reads: to create `analysis.pdf` the `analysis.Rmd` needs to be rendered, which depends on `data/clean.csv`. This dependency is itself a target. To create `data/clean.csv`, `R/clean.R` and `data/iris.csv` are needed. If you type `make analysis.pdf` `make` first checks whether the dependencies do exist and if not whether or it knows how to create them. So if `data/clean.csv` does not exist, `make` creates it. The same thing happens if one of the dependencies of a target is newer then the target itself, then updates everything that directly or indirectly depends on it. So if `data\iris.csv` is newer then `data\clean.csv`, make attempts to recreate `data\clean.csv` and `analysis.pdf`. If there is a dependency which is missing, and there is no recipe to make it, `make` stops with an error. It is a convention to have a target `all` which depends on everything and make it the first target in a `Makefile`, then the command `make` without any argument automatically creates everything. `Build All` from within RStudio also triggers this process.

```
all: analysis.pdf
```

![Build Pane in RStudio](Images/build-pane.png)

If you have followed the above workflow a fellow researcher is three commands away from fully reproducing your analysis:

```{bash, eval=FALSE}
git clone https://github.com/sirfisher/iris.git
cd iris
make all
```

However, this relies on the crucial assumption that your computational environments are sufficiently compatible e.g., everything needed is installed (R + Packages) at a compatible version of the software.
